Sequential rank agreement methods for comparison of ranked lists

### 2016-08-02 latex/superranker-manuscript.tex
Smårettelser

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify change-points in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

Other more recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. 
% One proposal is denoted the average overlap (AO) and used for
% comparison later on. 
Specifically, \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series to ensure that the top is
weighted higher than the potentially non-informative bottom of the
lists.  It is possible to use the existing methods to calculate
agreement of lists until a given depth, i.e., limited to the $d$ items
of each list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

@@

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could be essentially be used. Our proposed approach allows us to
compare more than two lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, places
high weight on items at the top of the list, accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
change-point, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored/partial lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker} which can be found
on CRAN.
% chunk 1

@@

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

@@

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the methods is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all depths $d$.  The sequential rank
agreement is equivalent to the pooled variance of the items found in
$S(d)$, and it can be interpreted as the average distance of the
individual rankings of the lists from the average ranking for each of
the items we have seen until depth $d$.  For an easier interpretation
we suggest to plot $\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding
to the pooled standard deviation.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify change-points in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using analytical asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

Other recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. 
% One proposal is denoted the average overlap (AO) and used for
% comparison later on. 
Specifically, \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series to ensure that the top is
weighted higher than the potentially non-informative bottom of the
lists.  It is possible to use the existing methods to calculate
agreement of lists until a given depth, i.e., limited to the $d$ items
of each list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

@@

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
change-point, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker} which can be found
on CRAN.
% chunk 1

@@

The agreement among the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

@@

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the rankings is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all depths $d$.  The sequential rank
agreement is equivalent to the pooled variance of the items found in
$S(d)$, and it can be interpreted as the average distance of the
individual rankings of the lists from the average ranking for each of
the items we have seen until depth $d$.  For an easier interpretation
we suggest to plot $\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding
to the pooled standard deviation.
